---
title: "Predicting ICU Mortality: A Classification Workflow in R"
description: "An end-to-end classification project in R predicting in-hospital mortality from vital signs recorded in the first hours of an ICU stay. Covers missing-value imputation with MICE, encoding of categorical variables, feature scaling, class weighting for imbalanced outcomes, and penalised logistic regression with ridge and lasso, evaluated by AUC."
author: "Andres Martinez Gonzalez"
date: "2021-01-16"
categories: [classification, machine learning, statistical learning, logistic regression, ridge, lasso, MICE, R]
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
---

## The problem

A patient arrives at an Intensive Care Unit. Within the first hours, monitors record a handful of vital signs: heart rate, blood pressure, respiratory rate, temperature, oxygen saturation, blood glucose. Each is summarised as a minimum, a maximum and a mean over that window.

The question is whether those early measurements carry enough signal to predict whether the patient will survive the hospital stay.

This is a binary classification problem, and a useful one to work through end to end, because the raw data has almost every complication a real dataset throws at you: missing values scattered across the clinical variables, categorical fields that need encoding, numerical features measured on wildly different scales, and — most importantly — a heavily imbalanced outcome, since most patients survive.

The plan is straightforward. Clean the data, engineer the features, split into training and validation sets, fit two penalised logistic regressions, and compare them on AUC. What follows is that sequence, with the reasoning for each decision spelled out.

## Setting up

Three things happen here. First, packages are loaded through a small routine that installs anything missing, so the document runs on any machine without manual setup. Second, the data is read straight from a public GitHub repository rather than a local path — the same reason: reproducibility. Third, two helper functions are defined up front.

```{r}
#| message: false
#| warning: false
#Install and load the packages needed 
list.of.packages <- c('caret', 'ggplot2','gsubfn' ,'FNN','kknn','fBasics' , 'mvtnorm' ,
                      'dplyr', 'mice' , 'fastDummies','glmnet', 'MLmetrics','ROCR','tidyr',
                      'tidyverse', 'magrittr', 'Hmisc')
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
sapply(list.of.packages, require, character.only = TRUE)
```

The training and test sets are read directly by URL.

```{r}
####Load the datasets  ####  
data_train<-read.csv('https://github.com/AndresMtnezGlez/heptaomicron/raw/main/train_prob_death.csv') # this is a relative path from current folder to the folder where you have the dataset
data_test<-read.csv('https://github.com/AndresMtnezGlez/heptaomicron/raw/main/test_prob_death.csv')
#head(data_train) #Peak the dataset first rows 
#head(data_test)  #Peak the dataset first rows 
#str(data_train)  #Check the structure of the variables
```

Two small helpers are worth defining once rather than repeating. `getmode()` returns the most frequent value of a vector, which is what we want when filling gaps in a categorical field — the mean is meaningless for a category, but the mode is a sensible default. `plot_miss()` draws the share of missing values in every column, which turns out to be the fastest way to see whether an imputation step actually worked.

Setting a seed makes everything below reproducible: the random split, the imputation draws, the sampling.

```{r}
set.seed(100)
getmode <- function(v){
  v=v[nchar(as.character(v))>0]
  uniqv <- unique(v)
  uniqv[which.max(tabulate(match(v, uniqv)))]
}

plot_miss <- function(data_train) {
 
  df <- as.data.frame(colSums(is.na(data_train))/nrow(data_train))
  df <- data.frame(row.names(df), df, row.names = NULL)
 
  ggplot(df, aes(x=row.names.df., y=colSums.is.na.data_train...nrow.data_train.)) + 
    geom_bar(stat = "identity",fill="#4199cb" )+
    ggtitle("Share of missing values per category")+
    coord_flip()
}
```

## Missing values

Before anything else, the two date fields are stored as character strings and need converting, otherwise any arithmetic on them silently fails.

```{r}
data_train$DOB <- as.Date(data_train$DOB)
data_train$ADMITTIME <- as.Date(data_train$ADMITTIME) 
str(data_train) #To check that it has worked
```

Now to the missing values. The first move is always to look at the extent of the problem rather than guess at it.

```{r}
####Missing values#### 
plot_miss(data_train)
```

The plot shows the gaps are concentrated in the clinical measurements — heart rate, blood pressure, glucose and so on — which makes sense: not every patient gets every measurement taken in their first hours.

The question is what to do about them. Dropping the affected rows would throw away a large share of the data. Filling each gap with the column mean is better, but it distorts the distribution: every imputed value sits exactly at the centre, which artificially shrinks the variance and weakens any correlation the variable has with the outcome.

The approach used here is **multiple imputation by chained equations** (MICE). Instead of filling a gap with a single summary statistic, MICE models each incomplete variable as a function of the others and draws a plausible value from that model. It does this iteratively — imputing one variable, then using the filled-in version to help impute the next, cycling until the values stabilise. The result preserves the relationships between variables, which is exactly what a downstream model needs.

The procedure starts from unconditional means (`meth="mean", maxit=0`) and then runs five iterations, generating three complete datasets.

```{r}
#| message: false
#| warning: false
#| results: hide
#Replace missing in categorical by most frecuent value 
l_num_features <- c('HeartRate_Min','HeartRate_Max','HeartRate_Mean','SysBP_Min','SysBP_Max','SysBP_Mean','DiasBP_Min','DiasBP_Max',
                  'DiasBP_Mean','MeanBP_Min','MeanBP_Max','MeanBP_Mean','RespRate_Min','RespRate_Max','RespRate_Mean','TempC_Min',
                  'TempC_Max','TempC_Mean','SpO2_Min','SpO2_Max','SpO2_Mean','Glucose_Min','Glucose_Max','Glucose_Mean')
df_num_feautes <- data_train[,l_num_features]
init <- mice(df_num_feautes, meth="mean", maxit=0)
imputation <- mice(df_num_feautes, method=init$method, maxit=5, m = 3)
```

One of those three imputed datasets is selected to complete the dataframe, and the missingness plot is redrawn to confirm the gaps are gone.

```{r}
data_train[,l_num_features]<-complete(imputation, 2)
plot_miss(data_train)
```

### Constructing age

The dataset gives a date of birth and an admission time, which is more useful as a single derived variable. Age at admission is a well-established predictor of ICU outcomes, so it is worth constructing explicitly rather than leaving the model to infer it from two dates.

```{r}
#Getting only the "year" from the dates and calculating the age 
data_train$DOB <- format(as.Date(data_train$DOB, format="%Y/%m/%d"),"%Y")
data_train$ADMITTIME <- format(as.Date(data_train$ADMITTIME, format="%Y/%m/%d"),"%Y")
data_train$DOB<- as.numeric(as.character(data_train$DOB))
data_train$ADMITTIME<- as.numeric(as.character(data_train$ADMITTIME))

data_train %<>% mutate(AGE = DOB - ADMITTIME)
data_train %<>% select(-DOB, -ADMITTIME)
```

## Categorical variables

Categorical fields cannot be fed to a regression as they stand. The obvious fix — assigning 1, 2, 3 to the categories — is wrong, because it imposes an ordering and a spacing that the data does not have: it tells the model that category 3 is somehow three times category 1, and that the step from 1 to 2 is the same size as the step from 2 to 3.

The correct encoding is a set of **dummy variables**, one binary column per category, which imposes no ordering at all.

The cost is dimensionality. A field with a hundred distinct values becomes a hundred columns, most of them almost entirely zeros, and the model loses tractability without gaining much. So the encoding here is deliberately selective: only two fields are dummified, gender and the first care unit the patient was admitted to. The reasoning for the second is clinical — units differ in the severity of the cases they take, so which unit a patient lands in carries information about their condition.

Identifiers and the remaining high-cardinality fields are dropped.

```{r}
#Dummify cateogorical variables of interest for probability of death
dta <- fastDummies::dummy_cols(data_train, select_columns =c("GENDER","FIRST_CAREUNIT" ))

dta %<>% select(-subject_id, -hadm_id,-icustay_id, -GENDER,-ADMISSION_TYPE,
                       -INSURANCE, -RELIGION, -MARITAL_STATUS, -ETHNICITY, -DIAGNOSIS, 
                       -ICD9_diagnosis, -FIRST_CAREUNIT)
```

## Scaling

The features are measured in units that have nothing to do with each other. Heart rate runs in the tens to low hundreds, temperature sits around 37, glucose can reach several hundred. A penalised regression shrinks coefficients toward zero, and the size of a coefficient depends on the scale of its variable — so without standardisation, the penalty falls unevenly across features for reasons that have nothing to do with how predictive they are.

`scale()` fixes this by centring each variable at zero and dividing by its standard deviation. The dummy variables and the outcome flag are excluded, since standardising a binary indicator serves no purpose.

```{r}
#| message: false
#| warning: false
#Standarize numerical variables (all but dummies and exitus flag)

l_num_var = names(dta)
l_no_num  = c('HOSPITAL_EXPIRE_FLAG','AGE',                 
                   "GENDER_F"             ,"GENDER_M"             ,"FIRST_CAREUNIT_CCU",  
                   "FIRST_CAREUNIT_CSRU"  ,"FIRST_CAREUNIT_MICU"  ,"FIRST_CAREUNIT_SICU", 
                   "FIRST_CAREUNIT_TSICU" )
l_num_var <- l_num_var[!(l_num_var %in% l_no_num)]

dta %<>%  mutate_at(l_num_var, funs(c(scale(.))))
```

A quick visual check that it worked: two variables that started on very different scales should now both be centred on zero with unit standard deviation.

```{r}
#Normalization worked 
aux <- select(dta,TempC_Min, Glucose_Min )#Let's sample two variables to see if 
df_tidy <- gather(aux, cols, value) 

ggplot(df_tidy, aes(x = value)) + 
  geom_density(aes(color=cols))+
  ggtitle("Density plot of normalize variables")
```

## Preparing the data for modelling

The outcome variable is recoded from a numeric flag to a factor with meaningful labels — `D` for death, `S` for survival. This matters for some of the modelling functions, which behave differently depending on whether the target is numeric or a factor.

```{r}
X <- select(dta,-'HOSPITAL_EXPIRE_FLAG')
y<-  select(dta,'HOSPITAL_EXPIRE_FLAG')
data <- cbind(X,y)

data$HOSPITAL_EXPIRE_FLAG<-as.character(data$HOSPITAL_EXPIRE_FLAG)
data$HOSPITAL_EXPIRE_FLAG[data$HOSPITAL_EXPIRE_FLAG=='1']<-'D' #1 now becomes "D" from Death
data$HOSPITAL_EXPIRE_FLAG[data$HOSPITAL_EXPIRE_FLAG!='D']<-'S' #0 now becomes "S" from Survival
data$HOSPITAL_EXPIRE_FLAG<-as.factor(data$HOSPITAL_EXPIRE_FLAG)
```

A test set has been supplied separately, but it comes without labels, so it cannot be used to measure performance. To get an honest estimate, the training data is split again: 95% to fit the model, 5% held back to evaluate it.

The distinction is worth stating clearly, since the naming is easy to confuse. `data_test` is the supplied, unlabelled set used for the final predictions. `datatest` is the held-out slice of the training data used for validation.

```{r}
sample_train <- sample(seq_len(nrow(data_train)), size = round(.95 * nrow(data_train)))
datatrain<-data[sample_train,]
Xtrain<-X[sample_train,]
ytrain<-y[sample_train,]
datatest<-data[-sample_train,]
Xtest<-X[-sample_train,]
ytest<-y[-sample_train,]
```

### Handling class imbalance

Now the central difficulty of this problem. Most patients survive, so the two classes are far from balanced.

This matters more than it first appears. A model that predicts *survival* for every single patient — that never once predicts a death — would still score a high accuracy, simply because it is right about the majority. It would also be completely useless, since the entire point is to identify the patients at risk.

Left alone, a fitting algorithm drifts toward exactly that behaviour, because the majority class dominates the loss function. The correction is to **weight** the observations: rarer outcomes are given more influence, so misclassifying a death costs the model more than misclassifying a survival.

```{r}
fraction_0 <- 1.5*rep(1 - sum(y == '0') / nrow(y), sum(y == '0'))
fraction_1 <- rep(1 - sum(y == '1') / nrow(y), sum(y == '1'))
weights <- numeric(nrow(y))
weighted <-TRUE
if (weighted == TRUE) {
  weights[y == '0'] <- fraction_0
  weights[y == '1'] <- fraction_1
} else {
  weights <- rep(1, nrow(y))
}
weights_train <- weights[sample_train]
weights_test <- weights[-sample_train]
```

The imbalance itself is worth seeing rather than just asserting.

```{r}
ggplot(y, aes(y=HOSPITAL_EXPIRE_FLAG, x="", colour = "#4199cb")) + geom_count() +
    ggtitle("Size of individuals with reasult death vs survivals") + 
    xlab("0 means survival, 1 means death")
```

## Modelling

Two penalised logistic regressions are fitted. Both start from the same idea: with this many features, an unpenalised logistic regression will overfit, chasing noise in the training data and generalising badly.

A penalty term discourages that by charging the model for the size of its coefficients. Variables that contribute little get shrunk toward zero, and the fit becomes simpler and more stable. This is **regularisation**, and the two standard forms differ in one important respect.

**Ridge** shrinks coefficients toward zero but never all the way — every variable stays in the model, just with reduced influence. This is what you want when domain knowledge says all the features are plausibly relevant.

**Lasso** can force coefficients to be exactly zero, dropping variables from the model entirely. It performs selection as well as shrinkage, which is what you want when you suspect many features are irrelevant.

Both are governed by a penalty parameter, lambda, which controls how aggressive the shrinkage is. Too small and the model overfits; too large and it underfits. The value is chosen by cross-validation: the data is split into folds, the model fitted repeatedly with different lambdas, and the value that gives the best out-of-fold performance is kept.

### Ridge regression

`cv.glmnet` with `alpha=0` selects ridge. The cross-validation curve plots the error against the log of lambda, with dashed vertical lines marking the value that minimises error and the largest value within one standard error of it.

```{r}
Ridge_fit=cv.glmnet(as.matrix(Xtrain),y=factor(ytrain),type.measure = 'auc',family='binomial',alpha=0,weights=weights_train)
plot(Ridge_fit)
```

The optimal lambda:

```{r}
Ridge_lambda = Ridge_fit$lambda.min
Ridge_fit$lambda.min
```

```{r}
summary(Ridge_fit)
```

### Evaluating with AUC

Accuracy is the wrong metric here, for the reason set out above: with imbalanced classes it rewards a model that ignores the minority.

**AUC** — the area under the ROC curve — is the appropriate alternative. It measures the model's ability to separate the two classes across every possible decision threshold, rather than at one arbitrary cutoff. It has a clean interpretation: the probability that a randomly chosen patient who died is assigned a higher risk score than a randomly chosen patient who survived. An AUC of 0.5 is a coin flip; 1.0 is perfect separation.

```{r}
preds = predict(Ridge_fit,newx = as.matrix(Xtest), type="response",s='lambda.min')
y_pred_acc<-as.factor(as.numeric(preds>0.5))# Compute different Evaluation metrics 
ytest<-as.factor(ytest)

# Prin and plot the AUC - ROC
pred <- prediction(preds,as.integer(ytest))
auc<-performance(pred,'auc')@y.values[[1]]
print(paste0('The AUC with Ridge Regression is:: ',auc))
```

The ridge model reaches an AUC of roughly 0.79. Against a baseline of 0.5 — which is what you would achieve by guessing — that is a substantial amount of signal extracted from measurements taken in the first hours of a stay.

The ROC curve itself shows the trade-off directly: each point is a threshold, and moving along the curve trades false positives against true positives.

```{r}
y_bin<-as.integer(ytest=='1')
auc<-performance(pred,'auc')@y.values[[1]]
perf <- performance(pred,"tpr","fpr")
plot(perf)
```

### Lasso regression

Same setup, `alpha=1` instead of `alpha=0`.

```{r}
Lasso_fit=cv.glmnet(as.matrix(Xtrain),y=factor(ytrain),family='binomial',alpha=1,type.measure = 'auc',weights = weights_train)
plot(Lasso_fit)
```

```{r}
preds = predict(Lasso_fit,newx = as.matrix(Xtest), type="response",s='lambda.min')
y_pred_acc<-as.factor(as.numeric(preds>0.5))# Compute different Evaluation metrics 
ytest<-as.factor(ytest)

y_bin<-as.integer(ytest=='1')

auc<-performance(pred,'auc')@y.values[[1]]
print(paste0('The AUC with Lasso Regression is:: ',auc))
```

```{r}
pred <- prediction(preds,y_bin)
perf <- performance(pred,"tpr","fpr")
plot(perf)
```

Lasso performs comparably but does not improve on ridge here. That is a mildly informative result in itself: if aggressive variable selection does not help, it suggests most of the clinical features are carrying at least some signal rather than a few doing all the work. Ridge is carried forward for the final predictions.

## Predictions on the held-out test set

The supplied test set has to go through **exactly** the same transformations as the training data — same date handling, same imputation, same dummy encoding, same scaling, same column ordering. Any discrepancy and the model will be fed features that do not mean what it learned them to mean.

This is the step where classification projects most often break silently, because a mismatch does not throw an error. It just produces bad predictions.

```{r}
#| message: false
#| warning: false
#| results: hide
data_test$DOB <- as.Date(data_test$DOB)
data_test$ADMITTIME <- as.Date(data_test$ADMITTIME)
l_num_features <- c('HeartRate_Min','HeartRate_Max','HeartRate_Mean','SysBP_Min','SysBP_Max','SysBP_Mean','DiasBP_Min','DiasBP_Max',
                  'DiasBP_Mean','MeanBP_Min','MeanBP_Max','MeanBP_Mean','RespRate_Min','RespRate_Max','RespRate_Mean','TempC_Min',
                  'TempC_Max','TempC_Mean','SpO2_Min','SpO2_Max','SpO2_Mean','Glucose_Min','Glucose_Max','Glucose_Mean')
df_num_feautes <- data_test[,l_num_features]
init <- mice(df_num_feautes, meth="mean", maxit=0)
```

```{r}
#| message: false
#| warning: false
#| results: hide
imputation <- mice(df_num_feautes, method=init$method, maxit=5, m = 3)
```

```{r}
data_test[,l_num_features]<-complete(imputation, 2)
plot_miss(data_test)
```

One column, `Glucose_Mean`, does not get imputed by MICE on the test set. Rather than leave the gap, it is filled with the column mean — a cruder fix, but adequate for a single variable.

```{r}
data_test$Glucose_Mean[is.na(data_test$Glucose_Mean)]<-mean(data_test$Glucose_Mean,na.rm=TRUE)
plot_miss(data_test)
```

Age is reconstructed as before.

```{r}
#Getting only the "year" from the dates and calculating the age 
data_test$DOB <- format(as.Date(data_test$DOB, format="%Y/%m/%d"),"%Y")
data_test$ADMITTIME <- format(as.Date(data_test$ADMITTIME, format="%Y/%m/%d"),"%Y")
data_test$DOB<- as.numeric(as.character(data_test$DOB))
data_test$ADMITTIME<- as.numeric(as.character(data_test$ADMITTIME))

data_test %<>% mutate(AGE = DOB - ADMITTIME)
data_test %<>% select(-DOB, -ADMITTIME)
```

The same two categorical fields are dummified. The ICU stay identifier is kept aside first, since it is needed to label the final predictions even though it must not enter the model.

```{r}
#Dummify cateogorical variables of interest for probability of death
dta <- fastDummies::dummy_cols(data_test, select_columns =c("GENDER","FIRST_CAREUNIT" ))

dta_test_ic <- select(dta, icustay_id)

dta %<>% select(-subject_id, -hadm_id,-icustay_id, -GENDER,-ADMISSION_TYPE,
                       -INSURANCE, -RELIGION, -MARITAL_STATUS, -ETHNICITY, -DIAGNOSIS, 
                       -ICD9_diagnosis, -FIRST_CAREUNIT)
```

And the numerical features are standardised on the same terms.

```{r}
#| message: false
#| warning: false
#Standarize numerical variables (all but dummies and exitus flag)

l_num_var = names(dta)
l_no_num  = c('HOSPITAL_EXPIRE_FLAG','AGE',                 
                   "GENDER_F"             ,"GENDER_M"             ,"FIRST_CAREUNIT_CCU",  
                   "FIRST_CAREUNIT_CSRU"  ,"FIRST_CAREUNIT_MICU"  ,"FIRST_CAREUNIT_SICU", 
                   "FIRST_CAREUNIT_TSICU" )
l_num_var <- l_num_var[!(l_num_var %in% l_no_num)]

dta %<>%  mutate_at(l_num_var, funs(c(scale(.))))
```

```{r}
dta_test <- dta
```

With the test set in the same shape as the training data, the fitted ridge model produces predicted probabilities, which are converted to class predictions at a 0.5 threshold.

```{r}
preds_test = predict(Ridge_fit,newx = as.matrix(dta_test), type="response",s=Ridge_lambda)
y_preds_test<-as.factor(as.numeric(preds_test>=0.5))# Compute different Evaluation metrics 
df_y_preds_test <- as.data.frame(y_preds_test)
```

Finally, predictions are joined back to their stay identifiers and written out.

```{r}
str(dta_test_ic)
```

```{r}
str(df_y_preds_test)
```

```{r}
names(dta_test_ic) <- "icustay_id"
names(df_y_preds_test) <- "HOSPITAL_EXPIRE_FLAG"
result_prob_death_AMG <- cbind(dta_test_ic,df_y_preds_test)
```

```{r}
#| eval: false
write.csv(result_prob_death_AMG,"result_prob_death_AMG.csv", row.names = FALSE)
```

## What this shows, and what it leaves out

The workflow scores 0.68 on the held-out competition set. The gap between that and the 0.79 measured on the internal validation split is itself instructive — it is the ordinary distance between an estimate computed on data drawn from the same pool as the training set and performance on genuinely unseen data.

A few things this treatment leaves on the table, in rough order of how much they would likely help:

- **Threshold selection.** The 0.5 cutoff is a default, not a choice. In a clinical setting the costs of a false negative and a false positive are not remotely equal, and the threshold should reflect that.
- **Nonlinear models.** Ridge and lasso are both linear in the features. Tree-based methods would capture interactions between vital signs without them having to be specified in advance.
- **Feature engineering.** Only minimum, maximum and mean are used. The *spread* between minimum and maximum — how unstable a patient's vitals are — is plausibly as informative as the level, and is not currently represented anywhere.
- **The discarded categorical fields.** Admission type and primary diagnosis were dropped for dimensionality reasons. Grouping them into a smaller number of clinically meaningful categories would make them usable.
